A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}

### 2023-11-04 paper.tex
Commit riportato dal 3 novembre - Revisore 2 - Augusto

\colorlet{punct}{red!60!black}
\definecolor{background}{HTML}{EEEEEE}
\definecolor{delim}{RGB}{20,105,176}
\colorlet{numb}{magenta!60!black}

\lstdefinelanguage{json}{
    basicstyle=\normalfont\ttfamily,
    numbers=left,
    numberstyle=\scriptsize,
    stepnumber=1,
    numbersep=8pt,
    showstringspaces=false,
    breaklines=true,
    frame=lines,
    backgroundcolor=\color{background},
    literate=
     *{0}{{{\color{numb}0}}}{1}
      {1}{{{\color{numb}1}}}{1}
      {2}{{{\color{numb}2}}}{1}
      {3}{{{\color{numb}3}}}{1}
      {4}{{{\color{numb}4}}}{1}
      {5}{{{\color{numb}5}}}{1}
      {6}{{{\color{numb}6}}}{1}
      {7}{{{\color{numb}7}}}{1}
      {8}{{{\color{numb}8}}}{1}
      {9}{{{\color{numb}9}}}{1}
      {:}{{{\color{punct}{:}}}}{1}
      {,}{{{\color{punct}{,}}}}{1}
      {\{}{{{\color{delim}{\{}}}}{1}
      {\}}{{{\color{delim}{\}}}}}{1}
      {[}{{{\color{delim}{[}}}}{1}
      {]}{{{\color{delim}{]}}}}{1},
}

@@

% Authors, for the paper (add full first names)
\Author{ Maria Grazia Deri $^{2,}$*\orcidC{}, Letizia Chiti $^{3}$\orcidA{}, Augusto Ciuffoletti $^{1}$\orcidB{} }

@@

Leisure activities play an important role in bringing profit to niche businesses. This creates a reinforcing effect that follows the negative dynamics seen above: a leisure activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact. Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a}.

@@

@@

In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity (a {\em dead zone} using the term used by Pearce and Gretzel \cite{pea12a}). We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.

@@

\begin{itemize}
	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance;
	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports;
	\item environmental impact: to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site; 
	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons;
	\item accessibility for stranger visitors: to limit the size of the plaque, the number of translations must be limited as well;
	\item update limits: to update the content, the plaque must be replaced;
	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation.
\end{itemize}

@@

A popular research direction covers the social aspects related to the use of the smartphone. W. Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered on the Internet. Such an assumption is in contrast with the title, which indicates a nature-based destination, where notably the Internet is not always reachable.

@@

@@

\begin{itemize}
	\item {\em Equitable use} is closely related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone,
	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
	\item{\em Tolerance for error}: there are no margins to use the device in a way that compromises user safety. The deliberate or accidental release of the passive device into the environment determines a minor pollution,
	\item{\em Low physical effort} holds, although the user needs to carry a smartphone,
	\item{\em Size and Space for approach and use} need to be carefully considered. In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort.
\end{itemize}

@@

There are two relevant trade-offs that the host needs to resolve. One is related to the visibility of the tag. The trade-off is between visual impact and visibility. The other is related to monitoring tag utilization. This may be useful for all sorts of planning activities and is discussed in section \ref{sec:results}.

@@

In recent years, the main tourist attractions have concentrated on the \emph{Lima} stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting, and sup. Other important tourist attractions are trekking and hiking, supported by a network of paths tracked by local CAI (\textit{Club Alpino Italiano}).

@@

\subsubsection{Tourism dynamics in \emph{Bagni di Lucca}, the \emph{Val di Lima} and\emph{Casoli}}

@@

Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways. According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities. Local stakeholders operating in the tourism system offer a variety of services all year round. Whenever possible the offer is enriched with original touristic approaches paying attention to sustainability. As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination. Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric.. The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.

@@

\begin{figure}
\subfloat[QR-tag parts]{
	\includegraphics[width = 0.55 \linewidth]{figure/QR-code-legend.png}
	\label{fig:qrcodelegend}
}
\hfill
\subfloat[QR-tag content on a 5.5' screen]{
	\includegraphics[width = 0.35 \linewidth]{figure/qr-code-content}
	\label{fig:qrcodecontent}
}
\caption{QR tag parts and content}
\end{figure}

\begin{figure}
\subfloat[Linked page: the map]{
	\includegraphics[width = 0.60 \linewidth]{figure/webpage-map}
	\label{fig:linkmap}
}
\subfloat[Linked page: survey form]{
	\includegraphics[width = 0.30 \linewidth]{figure/webpage-poll.png}
	\label{fig:linkpoll}
}
\caption{Content of the Web page linked to the URL in a QR tag}
\label{fig:webpage}
\end{figure}

\begin{table}
    \begin{tabular} {|>{\raggedright\arraybackslash}p{3cm}|l|l|l|l|}
        \hline 
        \textbf{Name} & \textbf{Long} & \textbf{Lat} & \textbf{URL Key} & \textbf{Length} \\ \hline
        Buca La Piella & 10.68361 & 44.03667 & t5ysrm & 476 \\ \hline
        Calendario celtico & 10.66985 & 44.03986 & my0kp8 & 484 \\ \hline
        Castello & 10.67062 & 44.04004 & 4l4r6y & 597 \\ \hline
        Chiesa dei SS. Donato e Andrea & 10.66947 & 44.03928 & lwtyx6 & 632 \\ \hline
        Iscrizione longobarda & 10.67244 & 44.03906 & 60m75s & 369 \\ \hline
        Lago di Casoli & 10.67761 & 44.03468 & xqjpbk & 369 \\ \hline
        Madonna di Castello & 10.66994 & 44.03960 & qlci89 & 395 \\ \hline
        Madonna di Col di Piano & 10.67760 & 44.03173 & 3w44wr & 414 \\ \hline
        Metato & 10.67764 & 44.03591 & sbgnl0 & 660 \\ \hline
        Sant'Andrea de Lacu & 10.67592 & 44.03530 & fxq83v & 657 \\ \hline
        Antro dell’Ugola & 10.68296 & 44.03871 & mprs0w & 226 \\ \hline
        Deviazione per Antro dell’Ugola & 10.68343 & 44.03971 & e4n2js & 88 \\ \hline
        Deviazione per Buca La Piella & 10.68427 & 44.03573 & hve4pj & 139 \\ \hline
        Deviazione per Sant’Andrea de Lacu & 10.67663 & 44.03507 & 13uav3 & 200 \\ \hline
        Ingresso Buca La Piella & 10.68358 & 44.03592 & 5ahvp8 & 190 \\ \hline
        Riparo sottoroccia lungo percorso Piella & 10.68401 & 44.03573 & kwr1wx & 286 \\    \hline
    \end{tabular}
\caption{The QR-tags placed around \emph{Casoli} . The URL key is a random code that appears on the web page linked to the QR code. The Length refers to the description in the QR tag and is in characters. \label{tab:qrtags}}
\end{table}

@@

The finalized master document contains a single \emph{GeoJSON} \emph{FeatureCollection} object containing one \emph{Feature} object for each tag. Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: its unique identifier \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text to be encoded in the QR code (see figure \ref{fig:geojson-sample}). The \emph{properties} are also shown on the UMap map.

@@

\begin{figure}
\subfloat[Editing a feature associated with a QR-tag in UMap]{
	\includegraphics[width = 0.35 \linewidth]{figure/umap-editing.png}
	\label{fig:umap-editing}
}
\hfill
\subfloat[\emph{GeoJSON} \emph{Object} representing the same the feature]{
	\includegraphics[width = 0.60 \linewidth]{figure/geojson-sample.png}
	\label{fig:geojson-sample}
}
\caption{Editing the \emph{GeoJSON} with uMap editor. The resulting file is automatically converted into printable QR-tags}
\label{fig:editing}
\end{figure}

@@

The proposed solution consists of a dozen QR tags placed at key locations. The total cost of the tags amounts to a few Euros for the printing service, although finding a shop with an appropriate printing device may be difficult. The impact on the environment is extremely limited, as shown in figure \ref{fig:TagPlacement}.

@@

Trading off cost and simplicity for effectiveness and measurability there are solutions, based, for instance, on Bluetooth communication or dedicated applications, that might solve such issues.

@@

The resulting implementation is original in the literature, and minimalistic in its deployment. The detailed description enables its reuse as a starting point in future initiatives.

>>>>

\colorlet{punct}{red!60!black}
\definecolor{background}{HTML}{EEEEEE}
\definecolor{delim}{RGB}{20,105,176}

\lstdefinelanguage{json}{
    basicstyle=\normalfont\ttfamily\footnotesize,
    numbers=left,
    numberstyle=\scriptsize,
    stepnumber=1,
    numbersep=8pt,
    showstringspaces=false,
    breaklines=true,
    frame=lines,
    backgroundcolor=\color{background}
}

@@

% Authors, for the paper (add full first names)
\Author{ Maria Grazia Deri $^{2,}$\orcidC{}, Letizia Chiti $^{3}$\orcidA{}, Augusto Ciuffoletti $^{1}$*\orcidB{} }

@@

Leisure activities play an important role in bringing profit to niche businesses. This creates a reinforcing effect that follows the negative dynamics seen above: a leisure activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact. Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.

@@

{\huge Anticipare qui uso dei QR code}

@@

In our study, we consider a range of solutions that support sustainable tourism, specifically the dynamic provision of information during a visit along a natural trail not covered by Internet connectivity. We consider the presence of two stakeholders for the signage itself \cite{wan22a}: the hosting organization (the {\em host} for short) which implements a signing installation, and the visitor (or {\em user} ) which extracts useful and enjoyable information from signing devices.

@@

\begin{itemize}
	\item dimensions: the board must be sufficiently large to contain the desired content, taking into account its readability from a distance;
	\item installation logistics: depending on the location, the transportation and placement of a plaque may require a basement or other sorts of supports;
	\item environmental impact: to be effective, the plaque has to be prominent, and this may negatively affect the quality of the site; 
	\item accessibility for the visually impaired: the plaque is not useful for visually impaired persons;
	\item accessibility for foreign visitors: to limit the size of the plaque, the number of translations must be limited as well;
	\item update limits: to update the content, the plaque must be replaced;
	\item removal logistics: when the board degrades it must be either removed or replaced, which entails waste disposal together with other issues similar to those found during the installation.
\end{itemize}

@@

A popular research direction covers the social aspects related to the use of the smartphones. W. Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is considered on the Internet. Such an assumption is in contrast with the title, which indicates a nature-based destination, where notably the Internet is not always reachable.

@@

Regarding the specific technology we refer to, QR codes have been invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015. In 2013 an early paper by Martinez et al. \cite{mar13a} suggests its utilization to implement a virtual tour in the {\em Las Quilamas Natural Park} in Spain. At that time the availability of QR-code readers was not as diffused as today, and therefore the work may be considered among the seminal ones. The authors proposed to use the QR code as a link to a panel in PDF format viewed on the user's portable device. A similar work in a urban environment is presented in \cite{fin13a} and \cite{tat15a}, the latter illustrating the QR system serving the monuments in the Serbian town of Niš.

Since then the use of the QR code technology for heritage promotion has been suggested in other papers. In \cite{sol18a} the authors describe its application in a museum: after defining the points of interest, tags have been placed and linked to related documents. The authors conducted a poll about user perception on two segments of users: Italians and Polish. The QR tags were not unanimously considered useful: in particular half of the Italian users considered them as not useful. It is interesting to note that the motivation of the other segment of users, less skeptical about usefulness, shared a solid educational motivation to access QR-linked content (70\% in the Polish group against 50\% in the Italian one). This fact sheds light on the necessity to provide content the user needs: curiosity is not a sufficient motivation.

Recent publications about the use of QR codes for heritage valorization do not show significant novelties. In 2023, the authors of \cite{pan23a} introduce the installation of six QR codes in a museum and analyze the reactions of 42 visitors. The tags are created using the Taplink \url{https://taplink.at/} web service and are linked to content on the Web.

There are examples of urban outdoor applications of QR tags, like

All the found publications share a common attitude, which can be summarized in the motto "analog portal to digital world", as in the title of \cite{bai12a}. None of them evaluates the possibility of storing text in the tag itself, thus eliminating the requirement of an Internet connection.

The present paper promotes the use of QR codes with sustainability and as the primary or only way of providing information to the visitor. In the literature, QR codes are considered ancillary to other forms of communication, like physical installations or the Web. Therefore, in our case the informational content of the tag, which is commonly ignored, becomes relevant. It is a new and original way of using QR codes for the delivery of information.

@@

\begin{itemize}
	\item {\em Equitable use} is closely related to the smartphone technology, which is itself considered a vehicle for equitability,
	\item {\em Flexibility in use} is enabled by the device capabilities, which allow listening instead of reading, translating the information in a different language, or storing it for later use,
	\item {\em Simple and intuitive use} holds since the operation requires a single application, possibly already installed since useful in many circumstances, and tag reading requires a single finger touch on the smartphone,
	\item{\em Perceptible information} is a critical requirement, which contrasts to keep low environmental intrusion. This point will be further discussed in the section devoted to the implementation,
	\item{\em Tolerance for error}: there is no chance of accidentally causing a hazardous situation using the QR tag. The unintended release of the passive device into the environment determines a minor pollution,
	\item{\em Low physical effort} holds, although the user needs to carry a smartphone,
	\item{\em Size and Space for approach and use} need to be carefully considered. In the case of the NFC tag the smartphone needs to be nearly in touch with the passive device, while the QR code must be in the line of sight and frameable without effort.
\end{itemize}

@@

The host needs to resolve two issues. One problem is the balance between visual impact and visibility. Monitoring tag utilization, which is useful for all types of planning activities, is the other topic. They are discussed in section \ref{sec:results}.

@@

In recent years, the main tourist attractions have concentrated on the \emph{Lima} stream, with some associations and private entities promoting outdoor experiences, particularly fluvial sports, such as canyoning, rafting, and paddleboarding. Other important tourist attractions are trekking and hiking, supported by a network of paths tracked by local CAI (\textit{Club Alpino Italiano}).

@@

\subsubsection{Tourism dynamics in \emph{Bagni di Lucca}, the \emph{Val di Lima} and \emph{Casoli}}

@@

Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways. According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities. Local stakeholders operating in the tourism system offer a variety of services all year round. Whenever possible the offer is enriched with original touristic approaches paying attention to sustainability. As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination. Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric. The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.

@@

\begin{figure}
\subfloat[QR-tag parts]{
	\includegraphics[width = 0.60 \linewidth]{figure/QR-code-legend.jpeg}
	\label{fig:qrcodelegend}
}
\hfill
\subfloat[QR-tag content on a 5.5' screen]{
	\includegraphics[width = 0.30 \linewidth]{figure/qr-code-content.jpg}
	\label{fig:qrcodecontent}
}
\caption{QR tag parts and content}
\end{figure}

%\begin{figure}
%\subfloat[Linked page: the map]{
%	\includegraphics[width = 0.60 \linewidth]{figure/webpage-map}
%	\label{fig:linkmap}
%}
%\subfloat[Editing a feature associated with a QR-tag in UMap]{
%	\includegraphics[width = 0.35 \linewidth]{figure/umap-editing.png}
%	\label{fig:umap-editing}
%}
%\subfloat[Linked page: survey form]{
%	\includegraphics[width = 0.30 \linewidth]{figure/webpage-poll.png}
%	\label{fig:linkpoll}
%}
%\caption{Content of the Web page linked to the URL in a QR tag}
%\label{fig:webpage}
%\end{figure}

\begin{figure}
\includegraphics[width =\linewidth]{figure/combo_gen.jpeg}
\caption{Generation of the QR tag and UMap visualization from UMap content. The uMap editor (bottom left) provides a simplified access to the GeoJSON description. The same interface allows to specify the template controlling the Web visualization of the map (right side). An ad-hoc application referenced in the paper allows to transform the uMap file downloaded from the server into a batch of printable QR tags (top left). The tag is highlighted in red in the picture on the Web display}
\end{figure}

\begin{table}
    \begin{tabular} {|>{\raggedright\arraybackslash}p{3cm}|l|l|l|l|}
        \hline 
        \textbf{Name} & \textbf{Long} & \textbf{Lat} & \textbf{FID} & \textbf{Length} \\ \hline
        Buca La Piella & 10.68361 & 44.03667 & t5ysrm & 476 \\ \hline
        Calendario celtico & 10.66985 & 44.03986 & my0kp8 & 484 \\ \hline
        Castello & 10.67062 & 44.04004 & 4l4r6y & 597 \\ \hline
        Chiesa dei SS. Donato e Andrea & 10.66947 & 44.03928 & lwtyx6 & 632 \\ \hline
        Iscrizione longobarda & 10.67244 & 44.03906 & 60m75s & 369 \\ \hline
        Lago di Casoli & 10.67761 & 44.03468 & xqjpbk & 369 \\ \hline
        Madonna di Castello & 10.66994 & 44.03960 & qlci89 & 395 \\ \hline
        Madonna di Col di Piano & 10.67760 & 44.03173 & 3w44wr & 414 \\ \hline
        Metato & 10.67764 & 44.03591 & sbgnl0 & 660 \\ \hline
        Sant'Andrea de Lacu & 10.67592 & 44.03530 & fxq83v & 657 \\ \hline
        Antro dell’Ugola & 10.68296 & 44.03871 & mprs0w & 226 \\ \hline
        Deviazione per Antro dell’Ugola & 10.68343 & 44.03971 & e4n2js & 88 \\ \hline
        Deviazione per Buca La Piella & 10.68427 & 44.03573 & hve4pj & 139 \\ \hline
        Deviazione per Sant’Andrea de Lacu & 10.67663 & 44.03507 & 13uav3 & 200 \\ \hline
        Ingresso Buca La Piella & 10.68358 & 44.03592 & 5ahvp8 & 190 \\ \hline
        Riparo sottoroccia lungo percorso Piella & 10.68401 & 44.03573 & kwr1wx & 286 \\    \hline
    \end{tabular}
\caption{The QR-tags placed around \emph{Casoli} . The FID (Feature IDentifier) is a random code that appears on the web page linked to the QR code. The Length refers to the description in the QR tag and is in characters. \label{tab:qrtags}}
\end{table}

@@

The finalized master document contains a single \emph{GeoJSON} \emph{FeatureCollection} object containing one \emph{Feature} object for each tag. Each \emph{Feature} contains a \emph{Geometry} object of type \emph{Point}, and a \emph{properties} object with fields containing information needed to create the tag: its unique identifier \emph{uid}, a random string of six lowercase letters used to produce the associated URL, the \emph{title} printed on the tag, and the text to be encoded in the QR code (see figure \ref{fig:geojson-sample})

\begin{table}
\begin{lstlisting}[language=json]
{
  "type": "Feature",
  "geometry": { 
    "type": "Point", 
    "coordinates": [10.67592,44.035298]
  },
  "properties": {
    "fid": "fxq83v",
    "name": "Sant'Andrea de Lacu",
    "description": "Secondo le ... ad uso abitativo.",
    "imageTag": "1m0Vad1Om2ihSVezPUkMKfuCUlg3ptC6z",
    "urlPrefix": "https://sites.google.com/ ..."
  }
}
\end{lstlisting}
\caption{\emph{GeoJSON} \emph{Object} of a feature associated with a QR-code. Ellipsis (...) are used to shorten long strings  \label{fig:geojson-sample}}
\end{table}

The \emph{properties} are also shown on the UMap map.

@@

%\begin{figure}
%\subfloat[Editing a feature associated with a QR-tag in UMap]{
%	\includegraphics[width = 0.35 \linewidth]{figure/umap-editing.png}
%	\label{fig:umap-editing}
%}
%\hfill
%\subfloat[\emph{GeoJSON} \emph{Object} representing the %same the feature]{
%	\includegraphics[width = 0.60 \linewidth]{figure/geojson-sample.png}
%	\label{fig:geojson-sample}
%}
%\caption{Editing the \emph{GeoJSON} with uMap editor. The resulting file is automatically converted into printable QR-tags}
%\label{fig:editing}
%\end{figure}

@@

The proposed solution consists of placing at selected locations the QR tags listed in table \ref{tab:qrtags}. The total cost of the tags amounts to a few Euros for the printing service, although finding a shop with an appropriate device may be difficult. The impact on the environment is extremely limited, as shown in figure \ref{fig:TagPlacement}.

@@

Trading off cost and simplicity for effectiveness and measurability there are alternative solutions.

One technology that fits the requirements of our task is Bluetooth Low Energy (BLE): the autonomy of a battery-operated BLE device is compatible with the periodic maintenance of other forms of signage. A Bluetooth device can send to the visitor's smartphone a beacon to indicate the presence of a passive sign, like a not-so-visible QR tag, or directly deliver information. However this requires the installation of the app on the visitor, and the design of the IT product including the device and the application, which may cost far more than the physical devices.

A different application-based solution privileging passive devices is based on a writable NFC tag close to the one delivering the content. To update the number of visitors that have read the informational NFC, the user application reads and increments the count recorded on the writable NFC. Here too the primary drawback compared with our QR-tag based solution is the need for an ad-hoc application. In addition the event of data pollution is not excluded.

The demonstration of the existence of such needs biases the technological evolution: for instance, if counting NFC reads were a frequent requirement, tags with such a feature would be produced at low costs, and the increment operation might become a default in NFC reading apps. The research on low-energy and low-cost devices is also that of demonstrating that the market for such devices is expanding.

@@

The resulting implementation is minimalistic in its deployment. The detailed description enables its reuse as a starting point in future initiatives.
